```{r libraries}
library(flip)
library(resamplingMCP)
library(ggplot2)
library(reshape2)
library(plyr)
library(dplyr)
library(magrittr)
```

```{r t vs permutatio test}

adaptive_ttest <- function(x){
    n <- length(x)
    n1 <- floor(n/2)
    p1 <- t.test(x[1:n1],alternative='greater')$p.value
    p2 <- t.test(x[(n1+1):n],alternative='greater')$p.value
    {sqrt(c(n1,n-n1)/n) * qnorm(c(p1,p2),lower=F)} %>% sum() %>% pnorm(lower=FALSE)
}

hetero_ttest <- function(x,s=2){
    n <- length(x)
    if(n %% s != 0) stop('Sample size has to be even')
    n1 <- floor(n/s)
    xs <- split(x,rep(1:s,each=n1))
    ps <- sapply(lapply(xs,t.test,alternative='greater'),`[[`,'p.value')
    {sqrt(n1/n) * qnorm(ps,lower=F)} %>% sum() %>% pnorm(lower=FALSE)
}

compare_tvsp <- function(n,B,alpha=.05,rdist=rnorm_hetero,...) {
    out <- sapply(1:B,function(i,n=n,...) {
        x <- rdist(n,...)
        c(power_ttest = t.test(x,alternative='greater')$p.value,
          power_ptest = flip(x,tail=1)@res$`p-value`,
          power_atest = hetero_ttest(x,s=2))
    },n=n,...)
    rowMeans(out<=alpha)
}

plist <- lapply(seq(0,5,.25),function(ncp) c(ncp=ncp,compare_tvsp(20,5000,ncp=ncp,df=1)))

power_comparison <- plist %>%
    do.call(what='rbind') %>%
        as.data.frame %>%
            melt(id.vars="ncp",variable.name='test',value.name='power')

do.call('rbind',plist)


pdf("/tmp/permutation_vs_ttest_hetero.pdf",width=7,height=4)
qplot(ncp,power,linetype=test,data=power_comparison,geom='line')
dev.off()

plist2 <- lapply(seq(0,2,.1),function(ncp) c(ncp=ncp,compare_tvsp(20,5000,rdist=rt,ncp=ncp,df=1)))

power_comparison <- plist2 %>%
    do.call(what='rbind') %>%
        as.data.frame %>%
            melt(id.vars="ncp",variable.name='test',value.name='power')

pdf("/tmp/permutation_vs_ttest_tdist.pdf",width=7,height=4)
qplot(ncp,power,linetype=test,data=power_comparison,geom='line')
dev.off()

plist3 <- lapply(seq(0,1,.05),function(ncp) c(ncp=ncp,compare_tvsp(20,5000,rdist=rnorm,mean=ncp)))

power_comparison <- plist3 %>%
    do.call(what='rbind') %>%
        as.data.frame %>%
            melt(id.vars="ncp",variable.name='test',value.name='power')

pdf("/tmp/permutation_vs_ttest_norm.pdf",width=7,height=4)
qplot(ncp,power,linetype=test,data=power_comparison,geom='line')
dev.off()

plist4 <- lapply(seq(0,1,.05),function(ncp) c(ncp=ncp,compare_tvsp(20,5000,ncp=ncp,df=4)))

power_comparison <- plist4 %>%
    do.call(what='rbind') %>%
        as.data.frame %>%
            melt(id.vars="ncp",variable.name='test',value.name='power')

pdf("/tmp/permutation_vs_ttest_t4.pdf",width=7,height=4)
qplot(ncp,power,linetype=test,data=power_comparison,geom='line')
dev.off()

library(parallel)
options(mc.cores=3)
plist5 <- mclapply(seq(0,3,.2),function(ncp) c(ncp=ncp,compare_tvsp(20,5000,rdist=rnorm_contaminated,mean=ncp,csigma=3)))

power_comparison <- plist5 %>%
    do.call(what='rbind') %>%
        as.data.frame %>%
            melt(id.vars="ncp",variable.name='test',value.name='power')

pdf("/tmp/permutation_vs_ttest_normcont.pdf",width=7,height=4)
ggplot(power_comparison)+geom_line(aes(ncp,power,linetype=test))
dev.off()




power_diff <- mutate(dcast(power_comparison,ncp~test),power_diff=power_ptest - power_ttest,power_rel=power_ptest/power_ttest-1)
qplot(ncp,power_diff,data=power_diff,geom='line')

out <- replicate(10000,adaptive_ttest(rt(10,df=1,ncp=1)))
out2 <- replicate(10000,flip(rt(10,df=1,ncp=1),alternative=1)@res$`p-value`)
out3 <- replicate(10000,hetero_ttest(rt(10,df=1,ncp=1),s=5))
out4 <- replicate(10000,hetero_ttest(rt(10,df=1,ncp=1),s=5))
out4 <- replicate(10000,hetero_ttest(rnorm(10,mean=0),s=5))
out4 <- replicate(10000,hetero_ttest(rnorm(10,df=1,ncp=0),s=5))
mean(out < .05)
mean(out2 < .05)
mean(out3 < .05)
mean(out4 < .05)


```


## Simulations under the null

### Randomization model

We draw samples from a number of distributions, and compute the expected conditional error rate, taking the expectation over all possible first stage randomizations. We then compute the expceted expected conditional error rate and the maximum expected conditional error rate.

```{r adaptive null}

rdist <- rnorm
MCMC  <- 100
library(parallel)
options(mc.cores=3)

type1_randomization <- function(rdist,n1,n,MCMC,B=1000,...){
    mclapply(1:MCMC,function(i){
        x <- rdist(n,...)
        g1 <- omega(rep(0:1,each=n1/2))
        list(pCER = mean(apply(g1,2,function(g1) permutation_CER(x[1:n1],g1,x[(n1+1):n],B=B))),
             normCER = mean(apply(g1,2,function(g1) normal_CER(x[1:n1],g1,n))),
             tCER = mean(apply(g1,2,function(g1) t_CER(x[1:n1],g1,n))))}) %>%
                 dplyr::bind_rows()
}

t1_norm_1 <- type1_randomization(rnorm,4,8,100)
t1_rt1_1 <- type1_randomization(rt,4,8,100,df=1)
t1_rt4_1 <- type1_randomization(rt,4,8,100,df=4)

t1_norm_1$distribution  <- "Normal"
t1_norm_1$sample_size  <- 8
t1_rt1_1$distribution <- "Student-t, df=1"
t1_rt1_1$sample_size <- 8
t1_rt4_1$distribution <- "Student-t, df=4"
t1_rt4_1$sample_size <- 8

t1_norm_2 <- type1_randomization(rnorm,6,12,100)
t1_rt1_2 <- type1_randomization(rt,6,12,100,df=1)
t1_rt4_2 <- type1_randomization(rt,6,12,100,df=4)

t1_norm_2$distribution  <- "Normal"
t1_norm_2$sample_size  <- 12
t1_rt1_2$distribution <- "Student-t, df=1"
t1_rt1_2$sample_size <- 12
t1_rt4_2$distribution <- "Student-t, df=4"
t1_rt4_2$sample_size <- 12

t1_norm_3 <- type1_randomization(rnorm,8,16,100,B=5000)
t1_rt1_3 <- type1_randomization(rt,8,16,100,B=5000,df=1)
t1_rt4_3 <- type1_randomization(rt,8,16,100,B=5000,df=4)

t1_norm_3$distribution  <- "Normal"
t1_norm_3$sample_size  <- 16
t1_rt1_3$distribution <- "Student-t, df=1"
t1_rt1_3$sample_size <- 16
t1_rt4_3$distribution <- "Student-t, df=4"
t1_rt4_3$sample_size <- 16

t1_norm_4 <- type1_randomization(rnorm,50,100,100,B=5000)
t1_rt1_4 <- type1_randomization(rt,50,100,100,B=5000,df=1)
t1_rt4_4 <- type1_randomization(rt,50,100,100,B=5000,df=4)

t1_norm_4$distribution  <- "Normal"
t1_norm_4$sample_size  <- 100
t1_rt1_4$distribution <- "Student-t, df=1"
t1_rt1_4$sample_size <- 100
t1_rt4_4$distribution <- "Student-t, df=4"
t1_rt4_4$sample_size <- 100

pdf('/tmp/box.pdf',width=8,height=5)
par(mfrow=c(1,2))
boxplot(t1_norm_2[,1:3])
boxplot(t1_norm_3[,1:3])
dev.off()

t1_sims <- bind_rows(list(t1_norm_1,t1_norm_2,t1_norm_3,t1_norm_4,
                          t1_rt1_1,t1_rt1_2,t1_rt1_3,t1_rt1_4,
                          t1_rt4_1,t1_rt4_2,t1_rt4_3,t1_rt4_4))

(t1_sims %>% melt(value.name="type_1",id.vars=c('distribution','sample_size'),variable.name="test") -> t1_sims_molten) %>% head

levels(t1_sims_molten$test) <- c("Permutation","z-test","t-test")

library(xtable)

t1_sims_molten %>%
    group_by(test,distribution,sample_size) %>%
        summarise(mean=mean(type_1),sd=sd(type_1)) %>%
            transmute(sample_size,type_1 = paste0("\"",round(mean,3)," (",round(sd,3),")\"")) %>%
                dcast(distribution+sample_size~test) %>%
                    xtable %>% print.xtable(include.rownames=FALSE)



save(t1_sims,file="data/t1_sims_1711.rda")

pdf('figures/t1_008.pdf')
boxplot(t1_norm_1)
dev.off()
pdf('figures/t1_012.pdf')
boxplot(t1_norm_2)
dev.off()
pdf('figures/t1_016.pdf')
boxplot(t1_norm_3)
dev.off()
pdf('figures/t1_016.pdf')
boxplot(t1_norm_4)
dev.off()








n1 <- 20
n <- 40
x <- rnorm(n)
g1 <- rep(0:1,each=n1/2)
system.time(cond_dist(x[1:n1],x[(n1+1):n],g1,g1,meandiff,100000))
system.time(sort(perm_dist(x[1:n1],x[(n1+1):n],g1,g1,meandiff,100000))[ceiling(.975*36)])


choose(4,2)^2

## Small sample
simulate_trials(1,4,8)
simulate_trials(1,6,12)

## Medium sample
simulate_trials(1,20,40)
simulate_trials(1,50,100)

## Large sample
simulate_trials(1,100,200)
simulate_trials(1,200,400)

```
### Conditional first

```{r adaptive null}
## Small sample
simulate_trials(1,4,8)
simulate_trials(1,6,12)

## Medium sample
simulate_trials(1,20,40)
simulate_trials(1,50,100)

## Large sample
simulate_trials(1,100,200)
simulate_trials(1,200,400)

```

## Conditional none

```{r adaptive null}
## Small sample
simulate_trials(1,4,8)
simulate_trials(1,6,12)

## Medium sample
simulate_trials(1,20,40)
simulate_trials(1,50,100)

## Large sample
simulate_trials(1,100,200)
simulate_trials(1,200,400)

```


```{r simple example}
x1 <- rnorm(12)/sqrt(2)
x2 <- rnorm(12)/sqrt(2)
g1 <- rep(0:1,each=6)
g2 <- rep(0:1,each=6)
x1 <- x1+g1
tstat(x1,g1)

z_test(x1,NULL,g1,NULL)
t_test(x1,NULL,g1,NULL)

o <- omega(g1,B=10)
sumdiff(x1,o)
meandiff(x1,o)
zstat(c(x1,x2),c(g1,g2))
tstat(c(x1,x2),c(g1,g2))
tstat(c(x1,x2),c(g1,g2))


normal_CER(x1,g1,length(x1)+length(x2))
t_CER(x1,g1,length(x1)+length(x2),)
boxplot(cbind(resamplingMCP:::perm_dist(x1,x2,g1,g2,meandiff,100),
              resamplingMCP:::cond_dist(x1,x2,g1,g2,meandiff,100)))
permutation_CER(x1,g1,x2,meandiff,B=10000)


```

```{r}
library(resamplingMCP)
library(parallel)
MCMC = 1000
B = 1000
options(mc.cores=min(detectCores()-1,30))

## Medium size trial
n  <- 12
n1 <- 6



## Type 1 error of design where second stage is as preplanned
## to run a 1 sample symulation, just
T1_preplanned <- type1(rcauchy,n=n,n1=n1,B=B,MCMC=MCMC,stat=meandiff)
colMeans(T1_preplanned)

# colMeans(T1_preplanned)
# ## Type 1 error of normal distribution CER
# T1_normal <- type1_normal_preplanned(n,n1,MCMC)
# colMeans(T1_normal)


## Type 1 error of design where second stage is as preplanned
# to run a 1 sample symulation, just
T1_preplanned <- type1_preplanned(n,n1,B,MCMC,one_sample=TRUE)
colMeans(T1_preplanned)
# preplanned influenced     normal 
# 0.00000000 0.00000000 0.02479908 


## Type 1 error of normal distribution CER
T1_normal <- type1_normal_preplanned(n,n1,MCMC,one_sample=TRUE)
colMeans(T1_normal)
# preplanned influenced 
# 0.02669741 0.02669741 

## Type 1 error of design where second stage is as preplanned, heteroscedastic errors

T1_preplanned_hetero <- type1_preplanned_hetero(n,n1,B,MCMC,one_sample=TRUE)
colMeans(T1_preplanned_hetero)
# preplanned influenced     normal 
# 0.0317010  0.0313780  0.3103739 


## Type 1 error of normal distribution CER, heteroscedastic errors
MCMC = 5000
B = 5000
T1_normal_hetero <- type1_normal_preplanned_hetero(n,n1,MCMC,B,one_sample=TRUE)
colMeans(T1_normal_hetero) 
# preplanned influenced 
# 0.2367942  0.2367942 

T1_tstat_hetero <- type1_tstat_hetero(n,n1,MCMC,one_sample=FALSE)
colMeans(T1_tstat_hetero)

# preplanned influenced 
# 0.2367942  0.2367942 




```
