---
title: "Simulation studies: Conditional permutation vs inverse normal combiantion tests"
subtitle: "Continuous data"
author: Florian Klinglmueller
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Vignette Title}
  \usepackage[utf8]{inputenc}
---


```{r}
library(magrittr)
library(plyr)
library(dplyr)
library(pander)
library(parallel)
library(bt88.03.704)
library(resamplingMCP)
options(mc.cores=detectCores()-1)
```

```{r comparison function,echo=F}

    

##' Adaptive permutation test one-sample problems
##'
##' @title One-sample adaptive permutation test
##' @template onesample_sims
##' @param combination_function Function to combine stage-wise (permutation) p-values
##' @param perms Maximum number of permutations to use when computing permutation p-values and conditional error rates
##' @author Florian Klinglmueller
##' @export
adaptive_permtest2_os <- function(x,n1,n,ne,test_statistic,perms=50000,resam=100,alpha=0.025){
    if(ne>n){
        xs <- split(x,rep(1:3,c(n1,n-n1,ne-n)))
        gs <- split(sign(x)>0,rep(1:3,c(n1,n-n1,ne-n)))
        A <- permutation_CER2(xs[[1]],gs[[1]],xs[[2]],xs[[3]],test_statistic,one_sample=TRUE,restricted=FALSE,
                              permutations=perms,subsamples=resam,alpha=alpha)
        q <- perm_test(xs[[2]],xs[[3]],gs[[2]],gs[[3]],test_statistic,restricted=FALSE,B=perms)
        return(A>=q)
    } else {
        xs <- split(x,rep(1:2,c(n1,ne-n1)))
        gs <- split(sign(x)>0,rep(1:2,c(n1,ne-n1)))
        return(alpha>=perm_test(xs[[1]],xs[[2]],gs[[1]],gs[[2]],test_statistic,restricted=FALSE,B=perms))
    }
}


compare_adaptive_tests <- function(n1,n,rule,rdist,test_statistic,mpfr=F,resam=100,perms=100,...){
    x <- rdist(n,...)
    ne <- rule(x[1:n1])
    if(ne>n){
        x <- c(x,rdist(ne-n,...))
    } else {
        ne <- n
    }
    list(ne = ne,
         permtest = adaptive_permtest_os(x,n1,n,ne,test_statistic,perms=perms),
         permtest2 = adaptive_permtest2_os(x,n1,n,ne,test_statistic,perms=perms,resam=resam),
##         ttest = adaptive_ttest_os(x,n1,n,ne,mpfr=mpfr),
         invnorm = adaptive_invnormtest_os(x,n1,n,ne))
##         npcomb = adaptive_npcombtest_os(x,n1,n,ne,test_statistic))
}

run_simulation <- function(n1,n,mean,sd,B=10,resam=100,perms=100,...){
    results <- mclapply2(1:B,function(i) compare_adaptive_tests(n1,n,cond_power_rule_norm,rnorm,diffmean,mean=mean,sd=sd,resam=resam,perms=perms,...))
    results %>% bind_rows %>% summarize(ASN=mean(ne),maxSN=max(ne),permtest2=mean(permtest2),invnorm=mean(invnorm))#,npcomb=mean(npcomb),permtest=mean(permtest))
}



```


## One-sample test procedures for adaptive designs

The package `resamplingMCP` implements a number of test procedures
that for adaptive designs the permit for a increase.

- Adaptive permutation test
- Adaptive non-parametric combination test
- Adaptive combiniation test using inverse normal combination function 
- Adaptive $t$-test


```{r examples}
n1 <- 6
n <- 12
ne <- 12
x <- rnorm(ne,1)
adaptive_permtest2_os(x,n1,n,ne,possum,resam=1000)

c(adaptive_permtest_os(x,n1,n,ne,possum),
  adaptive_permtest2_os(x,n1,n,ne,possum,resam=1000),
  adaptive_ttest_os(x,n1,n,ne),
  adaptive_invnormtest_os(x,n1,n,ne),
  adaptive_npcombtest_os(x,n1,n,ne,possum))

```
## Some quick simulation studies


```{r try comparison}
list(
## one example fixed rule
compare_adaptive_tests(6,12,function(x) 3,rnorm,diffmean,mean=.5,resam=1000),
## one example conditional power rule
compare_adaptive_tests(6,12,cond_power_rule_norm,rnorm,diffmean,mean=.5,resam=1000),
## one example conditional power rule
compare_adaptive_tests(6,12,cond_power_rule_t,rnorm,diffmean,mean=.5,resam=1000)) %>% bind_rows
```


## Let's do all of this in a more systematic way


### Small sample scenario

We start with a preplanned sample size of twelve. Which provides $6$
observations for each group. Such that we will start all trials with a
first stage of size $6$ and second stage of at least $6$.

### Normal distribution

First let's have a look at the normal distribution. This is what both
the adaptive $t$-test and the inverse normal combination (of
$t$-tests) test assume.

### Type 1 error 

We look at a couple of parameter settings. First we consider scenarios
under the null. The remaining parameters that are of most interest are
preplanned sample size and the standard deviation of observations. The
scenario is chosen such that if data were distributed with a mean of
$\mu_0 = 1$ and standard deviation $\sigma_0 = 1$ a sample size of 12
would give a power of roughly 88% for the one-sample $t$-test against
a one-sided alternative (computed using R's function `power.t.test`).

During the interim analysis the sample size for the second stage will
be reassessed using a conditional power rule. That is, the variance is
estimated using the observed first stage outcomes and the second stage
sample size is increased to provide a conditional power of $90\%$
using the standard normal sample size formula

$$\left[\frac{\mu_0}{s(\boldsymbol{x}^F)} \left(\Phi^{-1}(1-\alpha) + \Phi^{-1}(1-\beta)\right)\right],$$

where $s(\boldsymbol{x}^F)$ denotes the standard deviation estimated from the first
stage observations, $\alpha$ is set to $0.025$ and $\beta = .1$.

```{r test_simulation,eval=F,echo=F,results='hide'}
scenarios <- expand.grid(n1=6,
                         n=12,
                         mean=1,
                         sd=c(1.5),resam=c(100,1000,10000),perms=c(100,1000,10000))

lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=1000)))) %>%
    bind_rows ->
    test_sim
        
test_sim
system.time(sim <- run_simulation(6,12,1,1.5,1000,perms=10000,resam=50000))

```


### Under the alternative

> We observed that the adaptive $t$-test does not control the Type I
> error rate in any scenario. This is surprising as (unconditional -
> finite sample) Type I error control of this method has been proven
> theoretically and confirmed numerically. Implementation of the
> approach however proved to be numerically challanging; especially
> the conditional density of the sum of squares at the preplanned end
> of the trial. It seems as if these problems are exacerbated in with
> small samples. We will therefore contact the authors of the original
> paper if they are aware of a better implementation. For the moment
> we have removed the corresponding results. 

We stick with the preplanned sample size of $n=12$ with a first stage
of $n^F = 6$ and then look at the power of the different test
procedures for mean values ranging from $0$ to $1.2$ in steps of
$.3$ and a standard deviation of outcomes $\sigma \in \{1,1.5\}$.


```{r sim_norm_alt,eval=F}
scenarios  <- expand.grid(n1=6,
                          n=12,
                          mean = round(seq(0,1.2,.3),1),
                          sd = c(1,1.5))

lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=50000)))) %>%
    bind_rows ->
        sim_norm_alt

save(sim_norm_alt,file=vfile("../data/sim_norm_alt",ending="rda"))
```

```{r}
library(pander)
load("../data/sim_norm_alt.rda")
pander(sim_norm_alt,split.tables=100)
```


## In a larger sample

So far we have only looked at scenarios with a fairly limited sample
size. As it turns out the non-parametric combination test is not very
usefull in such a situation which is due to the discreteness of the
permutation distribution.

We now look at simulated trials with slightly larger sample sizes. We
set the preplanned sample size to $n = 100$ with a first stage of $n^F
= 50$. We simulate under the null hypothesis for standard deviations
of $2, 3, 9, 12$. 

### Under the alternative

Finally we end our investigation into the operating characteristics of
adaptive tests under normality with a look at their performance under
the alternative for a larger preplanned trial. Again we set the
preplanned total sample size to $n=100$ with a first stage of $n^F =
50$. Again we compare the power of the different test procedures for
mean values ranging from $.1$ to $1$ in steps of $.1$. The standard
deviation of outcomes is now fixed at $\sigma = 3$. This was chosen so
that the preplanned trial (assuming $\mu_0=1$, $\sigma_0=3$) has a
power of about $90%$ ($91%$ according to `power.t.test`).



```{r,eval=F}
scenarios <- expand.grid(n1=50,
                         n=100,
                         mean=round(seq(0,1.2,.3),2),
                         sd=c(3,4))
                         
library(parallel)
lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=1000)))) %>% bind_rows -> sim_norm_alt_ls

save(sim_norm_null,file=vfile("../data/sim_norm_alt_ls"))


```
## Non-normal data

```{r invnorm,eval=F,echo=F}
run_simulation_cont <- function(n1,n,mean,sd,cshift,csd,cprop,B=10,resam=100,perms=100,...){
    results <- mclapply2(1:B,function(i) compare_adaptive_tests(n1,n,cond_power_rule_norm,rnorm_cont,diffmean,mean=mean,sd=sd,cshift=cshift,csd=csd,cprop=cprop,resam=resam,perms=perms,...))
    results %>% bind_rows %>% summarize(ASN=mean(ne),maxSN=max(ne),permtest2=mean(permtest2),invnorm=mean(invnorm))#,npcomb=mean(npcomb),permtest=mean(permtest))
}

scenarios <- expand.grid(n1=6,
                         n=12,
                         mean=round(seq(0,1.2,.3),2),
                         sd=c(1,1.5),
                         cshift = 0,
                         csd = c(3,10),
                         cprop = c(.1,.3),
                         perms = 10000,
                         resam = 10000)

lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation_cont',c(scenarios[p,],B=1000)))) %>%
    bind_rows ->
    sim_normcont

save(sim_normcont,file=vfile("../data/sim_norm_cont"))

```
```{r invnorm_limit,eval=F,echo=F}

cond_power_rule_norm_triple <- function(x1){
    cond_power_rule_norm(x1,maxN=length(x1)*3)
}

run_simulation_cont_cap<- function(n1,n,mean,sd,cshift,csd,cprop,B=10,resam=100,perms=100,...){
    results <- mclapply2(1:B,function(i) compare_adaptive_tests(n1,n,cond_power_rule_norm_triple,rnorm_cont,signedranks,mean=mean,sd=sd,cshift=cshift,csd=csd,cprop=cprop,resam=resam,perms=perms,...))
    results %>% bind_rows %>% summarize(ASN=mean(ne),maxSN=max(ne),permtest2=mean(permtest2),invnorm=mean(invnorm))#,npcomb=mean(npcomb),permtest=mean(permtest))
}

scenarios <- expand.grid(n1=6,
                         n=12,
                         mean=round(seq(0,1.2,.3),2),
                         sd=c(1,1.5),
                         cshift = 0,
                         csd = c(3,10),
                         cprop = c(.1,.3),
                         perms = 10000,
                         resam = 10000)

simAlt <- run_simulation_cont_cap(n1=6,n=12,mean=1,sd=1.2,cshift=0,csd=3,cprop=.1,B=1000,perms=1000,resam=1000)
simNull <- run_simulation_cont_cap(n1=6,n=12,mean=0,sd=1.2,cshift=0,csd=3,cprop=.1,B=1000,perms=1000,resam=1000)

lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation_cont_cap',c(scenarios[p,],B=1000)))) %>%
    bind_rows ->
    sim_normcont_rk

save(sim_normcont_rk,file=vfile("../data/sim_norm_cont_rk"))

```

```{r master_simulation,eval=F,echo=F}
run_scenario <- function(B,...){
    ## just a wrapper around mclapply2
    results <- mclapply2(1:B,function(i) compare_adaptive_tests(...))
    results %>% bind_rows %>% summarize(ASN=mean(ne),maxSN=max(ne),permtest=mean(permtest),permtest2=mean(permtest2),invnorm=mean(invnorm))#,npcomb=mean(npcomb),permtest=mean(permtest))
}

run_simulation_general <- function(scenarios,B=100,...){
    params <- list(...)
    bind_rows(lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_scenario',c(B=B,scenarios[p,],params)))))
}
scenarios_test <- expand.grid(n1=50,
                              n=100,
                              mean=c(0,1),
                              sd=c(3,5),
                              perms = 500,
                              resam = 500)

quicktest <- run_simulation_general(scenarios_test,B=100,rule=cond_power_rule_norm,test_statistic=diffmean,rdist=rnorm)


scenarios_normal_small <- expand.grid(n1=6,
                                      n=12,
                                      mean=round(seq(0,1,.5),2),
                                      sd=c(1,1.2,1.5),
                                      perms = 10000,
                                      resam = 5000)

sim_normal_small <- run_simulation_general(scenarios_normal_small,B=5000,rule=cond_power_rule_norm,test_statistic=diffmean,rdist=rnorm)
save(scenarios_normal_small,file=vfile('~/repos/resamplingMCP/data/scenarios_normal_small'))



scenarios_normal_large <- expand.grid(n1=50,
                                      n=100,
                                      mean=round(seq(0,1,.5),2),
                                      sd=3:5,
                                      perms = 10000,
                                      resam = 5000)

sim_normal_large <- run_simulation_general(scenarios_normal_large,B=5000,rule=cond_power_rule_norm,test_statistic=diffmean,rdist=rnorm)
save(sim_normal_large,file=vfile("~/repos/resamplingMCP/data/sim_normal_large"))

scenarios_nonnormal_large <- expand.grid(n1=50,
                               n=100,
                               mean=round(seq(0,1.2,.4),2),
                               sd=3:5,
                               cshift = 0,
                               csd = c(10,30),
                               cprop = c(.1,.3),
                               perms = 10000,
                               resam = 5000)

sim_nonnormal_large <- run_simulation_general(scenarios_nonnormal_large,B=5000,rule=cond_power_rule_norm,test_statistic=diffmean,rdist=rnorm_cont)
save(sim_nonnormal_large,file=vfile("~/repos/resamplingMCP/data/sim_nonnormal_large"))


scenarios_nonnormal_small <- expand.grid(n1=6,
                               n=12,
                               ncp=round(seq(0,1,.5),2),
                               df=c(1,2,3),
                               perms = 10000,
                               resam = 5000)

cond_power_rule_norm_max <- function(x1) cond_power_rule_norm(x1,maxN=500)
sim_nonnormal_small <- run_simulation_general(scenarios_nonnormal_small,B=5000,rule=cond_power_rule_norm_max,test_statistic=diffmean,rdist=rnorm_hetero)
save(sim_nonnormal_small,file=vfile("~/repos/resamplingMCP/data/sim_nonnormal_small"))


```
