---
title: "Simulation studies on the robustness of adaptive tests"
subtitle: "Numerical results"
author: Florian Klinglmueller
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Vignette Title}
  \usepackage[utf8]{inputenc}
---


```{r}
library(resamplingMCP)
library(magrittr)
library(plyr)
library(dplyr)
library(pander)
library(parallel)
library(bt88.03.704)
options(mc.cores=detectCores()-1)


```
## A number of different adaptive procedures

```{r procs,echo=FALSE}

## interface

## permutation test


```
## Some examples

```{r examples}
n1 <- 6
n <- 12
ne <- 15
x <- rnorm(15,1)

c(adaptive_permtest_os(x,n1,n,ne,possum),
  adaptive_ttest_os(x,n1,n,ne),
  adaptive_invnormtest_os(x,n1,n,ne),
  adaptive_npcombtest_os(x,n1,n,ne,possum))

```
## Some quick simulation studies

```{r comparison function,echo=F}

run_simulation <- function(n1,n,mean,sd,B=10,...) mclapply2(1:B,function(i) compare_adaptive_tests(n1,n,cond_power_rule_norm,rnorm,diffmean,mean=mean,sd=sd)) %>% bind_rows %>% colMeans


```

```{r try comparison}
list(
## one example fixed rule
compare_adaptive_tests(6,12,function(x) 3,rnorm,diffmean,mean=.5),
## one example conditional power rule
compare_adaptive_tests(6,12,cond_power_rule_norm,rnorm,diffmean,mean=.5),
## one example conditional power rule
compare_adaptive_tests(6,12,cond_power_rule_t,rnorm,diffmean,mean=.5)) %>% bind_rows
```

```{r simulation example,eval=F}
## lets replicate
sim <- bind_rows(lapply(1:100,function(i) compare_adaptive_tests(10,20,cond_power_rule_norm,rnorm_scont,diffmean,mean=1,cmean=4,cprop=0.2,csd=.1)))
colMeans(sim)
```

## Let's do all of this in a more systematic way


### Small sample scenario

We start with a preplanned sample size of twelve. Which provides $6$
observations for each group. Such that we will start all trials with a
first stage of size $6$ and second stage of at least $6$.

### Normal distribution

First let's have a look at the normal distribution. This is what both
the adaptive $t$-test and the inverse normal combination (of
$t$-tests) test assume.

### Type 1 error 

We look at a couple of parameter settings. First we consider scenarios
under the null. The remaining parameters that are of most interest are
preplanned sample size and the standard deviation of observations. The
scenario is chosen such that if data were distributed with a mean of
$\mu_0 = 1$ and standard deviation $\sigma_0 = 1$ a sample size of 12
would give a power of roughly 88% for the one-sample $t$-test against
a one-sided alternative (computed using R's function `power.t.test`).

During the interim analysis the sample size for the second stage will
be reassessed using a conditional power rule. That is, the variance is
estimated using the observed first stage outcomes and the second stage
sample size is increased to provide a conditional power of $90\%$
using the standard normal sample size formula

$$\left[\frac{\mu_0}{s(\boldsymbol{x}^F)} \left(\Phi^{-1}(1-\alpha) + \Phi^{-1}(1-\beta)\right)\right],$$

where $s(\boldsymbol{x}^F)$ denotes the standard deviation estimated from the first
stage observations, $\alpha$ is set to $0.025$ and $\beta = .1$.


```{r sim_norm_null,eval=F}




scenarios <- expand.grid(n1=6,
                         n=12,
                         mean=0,
                         sd=c(.5,1,1.5,2))
                         
lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=50000)))) %>%
    bind_rows ->
    sim_norm_null

save(sim_norm_null,file="~/repos/resamplingMCP/data/sim_norm_null.rda")

```

```{r}
load("~/repos/resamplingMCP/data/sim_norm_null.rda")
pander(sim_norm_null,split.tables=100)
```
### Under the alternative

> Observe that the adaptive $t$-test does not control the Type I error
> rate in any scenario. This is surprising as (unconditional - finite
> sample) Type I error control of this method has been proven
> theoretically and confirmed numerically. Implementation of the
> approach however proved to be numerically challanging; especially the
> conditional density of the sum of squares at the preplanned end of the
> trial. It seems as if this problems are exacerbated in with small
> samples. We will therefore contact the authors of the original paper
> if they are aware of a better implementation.

We stick with the preplanned sample size of $n=12$ with a first stage
of $n^F = 6$ and then look at the power of the different test
procedures for mean values ranging from $.1$ to $1$ in steps of
$.1$. The standard deviation of outcomes is fixed at $\sigma = 1$.


```{r sim_norm_alt,eval=F}

scenarios  <- expand.grid(n1=6,
                          n=12,
                          mean = round(seq(.1,1,.1),1),
                          sd = 1)

lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=50000)))) %>%
    bind_rows ->
        sim_norm_alt

save(sim_norm_alt,file="../data/sim_norm_alt.rda")
```

```{r}
library(pander)
load("../data/sim_norm_alt.rda")
pander(sim_norm_alt,split.tables=100)
```


## In a larger sample

So far we have only looked at scenarios with a fairly limited sample
size. As it turns out the non-parametric combination test is not very
usefull in such a situation which is due to the discreteness of the
permutation distribution.

We now look at simulated trials with slightly larger sample sizes. We
set the preplanned sample size to $n = 100$ with a first stage of $n^F
= 50$. We simulate under the null hypothesis for standard deviations
of $2, 3, 9, 12$. 

```{r sim_norm_null_ls,eval=F}

scenarios <- expand.grid(n1=50,
                         n=100,
                         mean=0,
                         sd=c(2,3,9,12))
                         
library(parallel)
options(mc.cores=3)
lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=1000)))) %>% bind_rows -> sim_norm_null_ls

save(sim_norm_null,file="../data/sim_norm_null_ls.rda")

```
### Under the alternative

Finally we end our investigation into the operating characteristics of
adaptive tests under normality with a look at their performance under
the alternative for a larger preplanned trial. Again we set the
preplanned total sample size to $n=100$ with a first stage of $n^F =
50$. Again we compare the power of the different test procedures for
mean values ranging from $.1$ to $1$ in steps of $.1$. The standard
deviation of outcomes is now fixed at $\sigma = 3$. This was chosen so
that the preplanned trial (assuming $\mu_0=1$, $\sigma_0=3$) has a
power of about $90%$ ($91%$ according to `power.t.test`).



```{r,eval=F}
scenarios <- expand.grid(n1=50,
                         n=100,
                         mean=seq(0.1,1,.1),
                         sd=3)
                         
library(parallel)
options(mc.cores=3)
lapply(1:nrow(scenarios),function(p) c(scenarios[p,],do.call('run_simulation',c(scenarios[p,],B=1000)))) %>% bind_rows -> sim_norm_null_ls

save(sim_norm_null,file="../data/sim_norm_null_ls.rda")


```
