# Permutation of adjusted p-values

A major problem applying graph based procedures based on permutation
tests is that for larger numbers of hypotheses (>20) computation
becomes infeasible due to the exponential growth of the number of
intersection hypotheses that have to be tested. Computational
shortcuts for general weighting strategies defined by graphs are not
available and likely do not exist, even, for very simple classes of
graphical weighting procedures. The simple graphical approach that is
based on weighted Bonferroni tests does not exploit the correlation or
any other association between test statistics and therefore suffers
from conservativeness. 

In this paper we suggest an alternative approach that permits to use a
graphical weighting procedure in combination with a multivariate
permutation test procedure. This is achieved by using the adjusted
p-values from a graph based multiple testing procedure as the test
statistics in a step-down min-P type test. In this way we can exploit
potential postive associations between test statistics while keeping
the computational burden low. The adjusted p-values are computed based
on the Bonferroni based graphical approach and are therefore computed
in linear time.

In a first step we have to show that this controls the FWER.

Then we need to show that this procedure improves on the Bonferroni
based procedure.

We need to show that the relative importances - implied by the graph -
are adhered to by this procedure.

Finally compare it to a full closed test procedures based on
multivariate permutation.

This article is accompanied by the r-package resamplingMCP.

```{r libraries}
library(flip)
library(gMCP)
library(resamplingMCP)
library(mvtnorm)
```

```{r}
help(package='gMCP')

G <- fixedSequence(5)

R <- diag(1,5)
R[cbind(1:4,2:5)] <- 1/2
R[cbind(2:5,1:4)] <- 1/2

p <- pnorm(z,lower=F)

out <- flip(z[,,1])
flite <- function(Y){
    pp <- flip(Y,tail=1,perms=100)@res$`p-value`
    -log10(gMCP::gMCP(G,pp)@adjPValues)
}

rescale <- function(out){
    if(class(out[[1]])=='flip.object'){
        sapply(out,function(p) p@res$`p-value`)
    } else {
        sapply(out,function(p) 10^(-p))
    }
}

library(parallel)
options(mc.cores=3)
MCMC = 100
z <- replicate(MCMC,rmvnorm(30,sigma=R))
z1 <- replicate(MCMC,rmvnorm(30,mean=c(1/2,1/2,0,0,0),sigma=R))
z2 <- replicate(MCMC,rmvnorm(30,mean=c(1/2,0,0,0,0),sigma=R))

perm2 <- flip(z2[,,1],statTest=flite,tail=+1)
plot(perm2)
flip(z2[,,1],statTest=flite,tail=-1)
flip(z2[,,1],statTest=flite,tail=0)



## global Null
pout <- mclapply(1:MCMC,function(i) flite(z[,,i]))
out <- mclapply(1:MCMC,function(i) flip(z[,,i],statTest=flite,tail=1,perms=100))

## first false
pout2 <- mclapply(1:MCMC,function(i) flite(z1[,,i]))
out1 <- mclapply(1:MCMC,function(i) flip(z1[,,i],statTest=flite,tail=1,perms=100))
## first two false
pout2 <- mclapply(1:MCMC,function(i) flite(z2[,,i]))
out2 <- mclapply(1:MCMC,function(i) flip(z2[,,i],statTest=flite,tail=1,perms=100))

```
