---
title: "The surprising robustness of the inverse
normal combination test under non-normality"
subtitle: "Some numerical results"
author: Florian Klinglmueller
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Vignette Title}
  \usepackage[utf8]{inputenc}
---

During a recent investigation we found that the inverse normal
combination test of stage-wise $t$-tests performs surprisingly well if
the parent distribution is not normal. It is well known that the
$t$-test controls the type I error for a wide range of parent
distribution. This robustness, however, comes at the price of reduced
power. As we will show the inverse normal combination test of
stagewise t-tests does actually outperform many nonparametric tests
in terms of power when the parent distribution is non-normal.

```{r}
library(magrittr)
library(plyr)
library(dplyr)
library(pander)
library(parallel)
library(bt88.03.704)
library(resamplingMCP)
library(flip)
options(mc.cores=detectCores()-1)
##' Adaptive permutation test one-sample problems
##'
##' @title One-sample adaptive permutation test
##' @template onesample_sims
##' @param combination_function Function to combine stage-wise (permutation) p-values
##' @param perms Maximum number of permutations to use when computing permutation p-values and conditional error rates
##' @author Florian Klinglmueller
##' @export
adaptive_permtest2_os <- function(x,n1,n,ne,test_statistic,perms=50000,resam=100,alpha=0.025){
    if(ne>n){
        xs <- split(x,rep(1:3,c(n1,n-n1,ne-n)))
        gs <- split(sign(x)>0,rep(1:3,c(n1,n-n1,ne-n)))
        A <- permutation_CER2(xs[[1]],gs[[1]],xs[[2]],xs[[3]],test_statistic,one_sample=TRUE,restricted=FALSE,
                              permutations=perms,subsamples=resam,alpha=alpha)
        q <- perm_test(xs[[2]],xs[[3]],gs[[2]],gs[[3]],test_statistic,restricted=FALSE,B=perms)
        return(A>=q)
    } else {
        xs <- split(x,rep(1:2,c(n1,ne-n1)))
        gs <- split(sign(x)>0,rep(1:2,c(n1,ne-n1)))
        return(alpha>=perm_test(xs[[1]],xs[[2]],gs[[1]],gs[[2]],test_statistic,restricted=FALSE,B=perms))
    }
}

```

```{r basic code}
# mclapply2(1:1000,function(i) adaptive_permtest2_os(rnorm_cont(40,mean=.5,cshift=0,csd=3),10,20,40,diffmean,perms=10000,resam=10)) %>% unlist %>% mean
# mclapply2(1:1000,function(i) adaptive_permtest2_os(rnorm_cont(40,mean=.2,cshift=0,csd=3),20,40,40,signedranks,perms=10000,resam=10)) %>% unlist %>% mean
B=1000
n=40
data <- matrix(rnorm_cont(B*n,mean=0.5,cshift=0,csd=3),nrow=n)
c(permW=mean(flip(data,statTest="Wilcoxon",perms=10000,tail=1)@res$`p-value`<.025),
  permT=mean(flip(data,perms=10000,tail=1)@res$`p-value`<.025),
  apermW=mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],20,40,40,signedranks,perms=10000),mc.progress=F) %>% unlist %>% mean,
  apermT=mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],20,40,40,diffmean,perms=10000),mc.progress=F) %>% unlist %>% mean,
  invnorm=mclapply2(1:B,function(i) adaptive_invnormtest_os(data[,i],20,40,40),mc.progress=F) %>% unlist %>% mean,
  ttest=mclapply2(1:B,function(i) (t.test(data[,i],alternative='greater')$p.value < 0.025),mc.progress=F) %>% unlist %>% mean,
  wilcox=mclapply2(1:B,function(i) (wilcox.test(data[,i],alternative='greater')$p.value < 0.025),mc.progress=F) %>% unlist %>% mean) -> nonadaptive

```
Some comments the Wilcoxon test is actually much better that the
t-test and also than the invnorm test.

```{r adaptive setting}
mclapply2(1:B,function(i) adaptive_invnormtest_os(data[,i],15,30,40),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],15,30,40,diffmean,perms=10000),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],15,30,40,signedranks,perms=10000),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_invnormtest_os(data[,i],10,20,40),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],10,20,40,diffmean,perms=10000),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_permtest_os(data[,i],10,20,40,signedranks,perms=10000),mc.progress=F) %>% unlist %>% mean
```

In the adaptive setting this
changes. Here the adaptive wilcoxon test looses quite a bit of power
when the preplanned second stage size is small in comparison to the
actual second stage sample size, such that the inverse normal
combination test may be favourable at a certain stage.


## Possible improvements of the adaptive signedrank test

One may however improve the adaptive signedrank test in two
ways.

### 1. Subsampling the extended second stage


This is already implemented in the package and has been shown to
provide improvements in power.

```{r subsampling}
mclapply2(1:B,function(i) adaptive_permtest2_os(data[,i],15,30,40,signedranks,perms=10000),mc.progress=F) %>% unlist %>% mean
mclapply2(1:B,function(i) adaptive_permtest2_os(data[,i],10,20,40,signedranks,perms=10000),mc.progress=F) %>% unlist %>% mean
```
The subsampling approach may be further improved using a
bootstrap procedure that also takes into account samples from the
first stage. The latter approach may be problematic as this introduces
dependeny between first and second stage data.


2. and second by looking at the conditional distribution of the
second stage ranks. Here the idea is to draw ranks for the second
stage by keeping the order of the first stage ranks as well as the
observed first stage signs.


See below for one stap of the sampling process for the second stage
data.

We need to assume that the probability of the second stage data to
fall between any of the first stage observations is the same. This
obviously does not hold for all parent distributions! A way out may be
bootstrap?


```{r}
x1 <- rnorm(10)
r1 <- rank(x1)

## preplanned second stage 10
## sample one possible second stage outcome
r2 <- sample(100,10)
r2[r2 %in% r1*10] <- r2[r2 %in% r1*10]+1
r2 <- rank(c(r2,r1*10))[1:10]

```
